---
title: 'PPM: Exploring 1^st^ order effects'
author: "Manny Gimond"
output:
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, small.mar=TRUE)

knitr::knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar = c(3, 3, 1, .1))  
})

```

------------------------------------------------------------------------

[Data for this tutorial can be downloaded [from here](./data.zip). Don't forget to unzip the files to a dedicated folder on your computer.]{.block1}

-----

> Don't forget to set the R session to the project folder via *Session \>\> Set Working Directory \>\> Choose Directory*.

-----

## Loading the data into R

First, we'll load the point pattern dataset, then we'll load two different raster datasets that will represent two separate 1^st^  order effects we suspect may help explain the distribution of stores within the study extent.

```{r results='hide', warning=FALSE, message=FALSE}
# Load packages
library(spatstat)
library(sf)
library(terra)

# Read state polygon data
s    <- st_read("MA.shp")
w    <- as.owin(s)
w.km <- rescale.owin(w, 1000)

# Read Walmart point data
s    <- st_read("Walmarts.shp")
p    <- as.ppp(s)
p.km <- rescale.ppp(p, 1000)
marks(p.km)  <- NULL
Window(p.km) <- w.km

# Read population density raster
img     <- rast("pop_sqmile.tif")
df      <- as.data.frame(img, xy = TRUE)
pop     <- as.im(df)
pop.km  <- rescale.im(pop, 1000)

# Read median income raster
img      <- rast("median_income.tif")
df       <- as.data.frame(img, xy = TRUE)
inc      <- as.im(df)
inc.km   <- rescale.im(inc, 1000)
```

Note that later in the script, we will make use of the `effectfun()` that requires that the raster layer be stored as a double data type. You can tell if your raster is stored as a double by typing `typeof(inc.km[])`, for example. If it's not stored as a double, you can force it to a double using the `as.double()` function (e.g. `inc[] <- as.double(inc[])`). 

Let's plot each raster with the Walmart point overlay. We'll use `R`'s base plotting environment.

```{r fig.height=3, small.mar = TRUE, fig.show='hold', echo = 2:7}
OP <- par(mfrow = c(1,2), mar=c(2,0,2,0)) # This creates a two-pane plotting window

plot(pop.km, ribside="bottom", main="Population")
plot(p.km, pch = 20, col=rgb(1,1,1,0.5), add=TRUE)

plot(inc.km, ribside="bottom", main="Income")
plot(p.km, pch = 20, col=rgb(1,1,1,0.6), add=TRUE)

par(OP)
```

## Modeling the Covariate Effect

We will first develop two different models that we think might define the Walmart's point intensity. Recall that *intensity* is a property of the underlying "process" (one being defined by the distribution of population density and the other being defined by the distribution of income)--this differs from *density* which is associated with the observed pattern. These models are represented in our R script by two raster layers: a population raster and an income raster. These models will be  **alternate** models that we'll denote as `Mpop` and `Minc` in the script. The models' structure will follow the form of a logistics model, but note that the models can take on many different forms and different levels of complexity.

We'll also create the null model, `Mo`, which will assume a **spatially uniform** (homogeneous) covariate. In other words, `Mo` will define the model where the intensity of the process is assumed to be the same across the entire study extent.

```{r}
Mpop <- ppm(p.km ~ pop.km) # Population model
Minc <- ppm(p.km ~ inc.km) # Income model
Mo   <- ppm(p.km ~ 1)      # Null model
```

Let's explore the model parameters. First, we'll look at `Mpop`.

```{r}
Mpop
```

The values of interest are the **intercept** (whose value is around `-6.3`) and the **coefficient** `pop.km` (whose value is around `0.0001`). Using these values, we can construct the mathematical relationship (noting that we are using the population raster and not the original population raster values):

$$
Walmart\ intensity(i)= e^{−6.3 + 0.0001\ population\ density_i}
$$ The above equation can be interpreted as follows: if the population density is `0`, then the Walmart intensity is $e^{−6.3}$ which is very close to `0`. So for every unit increase of the population density (i.e. one person per square mile), there is an increase in Walmart intensity by a factor of $e^{0.0001}$ stores per square kilometer.

Likewise, we can extract the parameters from the `Minc` model and construct its equation.

```{r}
Minc
```

$$
Walmart\ intensity(i) = e^{−5.58\ −1.66e^{−5}\ Income(i)}
$$ Note the negative (decreasing) relationship between income distribution and Walmart density.

Next, we'll extract the null model results:

```{r}
Mo
```

This gives us the following equation for the homogeneous process:

$$
Walmart\ intensity(i) = e^{−6.17} = 0.00209
$$

which is nothing more than the number of stores per unit area (44 stores / 21,000km^2^ = 0.00209).

### Plotting the competing models

```{r fig.show='hold', fig.height=3, small.mar =TRUE}
OP <- par(mfrow = c(1,2), mar = c(4,4,2,1))  # This creates a two-pane plotting window

plot(effectfun(Mpop, "pop.km", se.fit = TRUE), main = "Population",
     ylab = "Walmarts per km2", xlab = "Population density", legend = FALSE)

plot(effectfun(Minc, "inc.km", se.fit = TRUE), main = "Income",
     ylab = "Walmarts per km2", xlab = "Income", legend = FALSE)

par(OP) # This reverts our plot window back to a one-panel window
```

Note the difference in relationships between the two models. In the first plot, we note a slightly increasing relationship between Walmart intensity and population density; this is to be expected since you would not expect to see Walmart stores in underpopulated areas. In the second plot, we note an inverse relationship between Walmart intensity and income---i.e. as an area's income increases, the Walmart intensity decreases.

The grey envelopes encompass the 95% confidence interval; i.e. the true estimate (black line) can fall anywhere within this envelope. Note how the envelope broadens near the upper end of the population density values--this suggests wide uncertainty in the estimated model.

To assess how well the above models explain the relationship between covariate and Walmart intensity, we will turn to hypothesis testing.

----------------

## Testing the Covariate Effect

To formally assess whether the covariate-based model provides a significantly better fit than the null model, we need a way to reduce that comparison to a single number. That number is called the **likelihood‑ratio test (LRT)  statistic**. 

* A model that fits the data better gives a **higher likelihood** for the observed pattern.
* The LRT statistic measures how much better the more complex model `Mpop` or `Minc` fit the data than the simpler "null" model `Mo`.

Think of the LRT as a kind of **"distance"** between the two models’ explanations of the same point pattern.

* Small LRT statistic -> the alternative model explains the pattern almost equally well
* Large LRT statictic -> the alternative model explains the pattern noticeably better

### The `Mpop` model test

We'll first apply a LRT test for the `Mpop` model.

```{r}
anova(Mo, Mpop, test = "LRT") # Compare null to population model
```

In `spatstat`, the LRT statistic appears as the difference in **deviance** in the output. We compare this value to a **reference distribution**--a picture of how big the LRT statistic (i.e. the difference-in-deviance value in the output) would usually be if the null model really were adequate. The following figure shows this distribution with the observed statistic displayed as a veritcial blue line.

```{r fig.width=5, fig.height= 3, echo = FALSE}
OP <- par(mar=c(3.9,3.9,1,1))
observed_deviance <- 4.2968
df_value <- 1

# Generate a sequence of values for the x-axis (LRT statistic range)
x <- seq(0, 10, length.out = 500)

# Calculate the density of the Chi-Squared distribution with 1 degree of freedom
y <- dchisq(x, df = df_value)

# Plot the analytical null distribution
plot(x, y, type = "l", lwd = 2, col = "grey30",
#     main = "Analytical Null Distribution (Chi-Squared, df=1)",
     xlab = "LRT Statistic (Difference in Deviance)",
     ylab = "Probability Density",
     ylim = c(0, 1)) # Limit y-axis for better visibility near the tail

# Add a vertical line at the observed Deviance value
abline(v = observed_deviance, col = "blue", lwd = 2, lty = 2)

# Shading the area representing the p-value (upper tail)
x_tail <- seq(observed_deviance, 10, length.out = 100)
y_tail <- dchisq(x_tail, df = df_value)
polygon(c(observed_deviance, x_tail, 10), c(0, y_tail, 0), col = rgb(1, 0, 0, 0.2), border = NA)
par(OP)
```

The reference distribution addresses the question **"If the null model were true, how likely is it that random noise alone would produce an LRT statistic (difference-in-deviance) this large (4.2968) or larger?**. The probability can be extracted from the plot or from the `anova()` output under the `pr` heading--this is the **p-value**. For the `Mpop` model, the p-value is **0.03819** or **3.8%**.

This means that if the population density **had no real effect**, then a statistic as large as 4.2968 would occur only about **3.8%** of the time. Because this is fairly unlikely, we interpret the result as evidence that including population density does help explain where points occur.

### The `Minc` model test

```{r}
anova(Mo, Minc, test = "LRT") # Compare null to income model
```

The model using median income (`Minc`) shows a very different result. The LRT statistic for `Minc` is **1.3388**, yielding a **p‑value of 0.2473**--roughly 25%. This value is **well within the range of what we would expect just by chance** if the null model (constant intensity) were true. A statistic this small is not unusual under the null model, meaning that including income as a covariate does not provide strong evidence of improvement. From this, we can conclude that income does not significantly explain the observed point distribution, at least not in the way population density does. Together, these results suggest that population density is a more relevant predictor of point intensity than income in this dataset.

### Additional Guidance on Interpreting p‑values

Finally, it’s important not to treat the **0.05 cutoff** as a strict “significant / not significant” decision rule. The value 0.05 has no theoretical justification, it is simply a historical convention—and relying on it too rigidly can lead to oversimplified or misleading interpretations. Instead, interpret the p‑value as a continuous measure of how surprising your data would be if the null model were true. A smaller p‑value means the observed improvement is less likely to be due to chance alone, while a larger p‑value means the improvement is more consistent with random variation. This perspective helps you think in terms of evidence and uncertainty, rather than treating statistical testing as a binary yes/no decision.
